Describing the scale and composition of calls for service in Detroit

### 2021-08-24 paper.Rmd
officer initiated change, writing

---
title: "Describing the scale and composition of police demand in Detroit."
author: "Samuel Langton (Netherlands Institute for the Study of Crime and Law Enforcement)"
output: pdf_document
header-includes:
  - \usepackage{lscape}
---

@@

There is a common misconception that police spend most of their time responding to and dealing with crime. Instead, the evidence suggests that public demand for police services originates from a diverse array of incidents -- many of which involve vulnerable people and require specific training and numerous

## Police demand

Estimating the scale of demand of police services is not a simple task, particularly when it comes to complex sources of demand, such as those originating from persons with mental ill-health. One common method has been to use emergency calls for service data. For instance, we can say _how many_ emergency calls were classified as involving mental ill-health in any given time period. We can take this one step further by calculating the _amount of time_ police spend dealing with such incidents. We can then make claims such as: "Police officers spend x% of their time responding to and resolving incidents involving mental ill-health".

Gathering such evidence -- which simply _describes_ the composition of public demand for the police -- is important for a number of reasons. Among them, it can help guide training to ensure that officers are suitably prepared for their duties, maximising well-being outcomes for the public and mitigating against unnecessary strain on officers. Understanding police demand can also contribute to how funds from central and local governments are decided and distributed to police forces and other public services. This has become particularly pertinent following recent 'defund the police' campaigns. With police performing a variety of non-crime-related duties, any changes to their role will likely have a knock-on effect to other sectors (and vice-versa).

\newpage
```{r fig1, fig.align = "center", fig.cap = "Proportional breakdown of deployed time spent on each incident category.", out.width = "\\textwidth"}
knitr::include_graphics("visuals/fig1_time_port.png")
```
\newpage

@@

## Conclusion

>>>>

---
title: "Describing the scale and composition of calls for service in Detroit"
author: "Samuel Langton (Netherlands Institute for the Study of Crime and Law Enforcement)"
bibliography: refs.bib
output:
  # word_document: default
  pdf_document: default
header-includes: \usepackage{lscape}
csl: apa.csl
abstract: In the spirit of a recent contribution by Ratcliffe (2021) published in
  _Crime Science_ this short paper describes the scale and composition of public demand
  for police services in Detroit, United States, during 2019. Findings broadly confirm
  those by Ratcliffe, namely, that a considerable proportion of police time is spent
  resolving a diverse array of (non-criminal) calls for service, including public
  health incidents. The temporal and spatial patterning of different types of public
  demand for police services are often distinct. The data and code to replicate analyses
  are made openly available.
---

Keywords: police, calls for service, 911, demand, reactive.

@@

Amidst austerity measures, growing public expectation, and scrutiny, understanding the public demand for police services has become a priority among evidence-based policing researchers and practitioners [@boulton2017calls]. Without a grasp on the scale ('how much?') and composition ('what type?') of police demand, we are likely to observe sub-optimal and inequitable outcomes for the public, the inappropriate distribution of public funds, and unnecessary strain on officers [@ellison2021understanding; @lum2021can]. Understanding the characteristics public demand for the police has become particularly pertinent following recent calls to rethink, and some cases, radically reform, the role and reach of contemporary police forces [@lum2021can].

A study from @ratcliffe2021policing in _Crime Science_ sought to describe the complexity and diversity of public demand for police services in Philadelphia, United States. The study was motivated by the recognition that police spend a considerable proportion of deployed resource resolving incidents which fall outside the traditional crime-fighting role of the police. Instead, the origins of public demand for the police can often be attributed to a lack of supply and/or accessibility failure in other organisations, such as those providing support for people requiring (mental) health assistance [@wood2021can; @van2017law]. The study's findings provided data-driven evidence to suggest that the police indeed spend considerable amounts of time dealing with incidents that could be resolved by other -- often more appropriate -- agencies. Comparable findings were subsequently reported in the United States across multiple (anonymised) police jurisdictions [@lum2021can]. In advance of recent contributions, this short paper aims to describe the public demand for police services in Detroit, United States, using open data and open code which can be used to reproduce the findings (see xxxxxxxxxxxxxxxxxxxxxxxxxxxx).

## Measuring demand

Using computer aided dispatch (CAD) data from the Philadelphia Police Department (PPD) the @ratcliffe2021policing study provided a breakdown of calls for service in a typical year, defined as 2019 (pre-COVID-19). Incident descriptions (e.g., investigate person) were classified into general categories, namely, community issues, crime, medical/public health, proactive policing, quality of life, and traffic duties. Incidents were subset for those originating from the public, and thus not initiated by an officer. Recognising that frequency counts of calls for service do not necessarily reflect the amount of police resource consumed by each incident, raw counts were supplemented by the time consumed by police officers dealing with each incident type. Findings indicated that around 55% of officer shift activity time was spent on crime incidents, with the remainder of time allocated to incidents involving medical/public health (9%), the community (7%), proactive policing (5%), quality of life (14%) and traffic duties (11%). Focusing on medical/public health incidents, findings indicated that calls concentrated in particular areas of the city and during particular hours of the day.

The data used for this study covers the city of Detroit in Michigan, United States. The City of Detroit publish comparable calls for service data through their open data portal.^[Accessible via https://data.detroitmi.gov/]. The raw data includes both citizen-initiated 911 calls to request police services and officer-initiated calls spanning back to September 2016. In alignment with @ratcliffe2021policing, the data used here is subset for the year 2019 and excludes those calls initiated by an officer. Each incident has supplementary information which describes the type of incident, the latitude and longitude coordinates, and a timestamp. For each incident, the response time and time on the scene are reported. In sum, these comprise the total time officers spent on the incident.

\newpage
```{r fig1, fig.align = "center", fig.cap = "Proportional breakdown of deployed time spent on each incident category. Graphic inspired by Ratcliffe (2012).", out.width = "\\textwidth"}
knitr::include_graphics("visuals/fig1_time_port.png")
```
\newpage

@@

## Discussion
